BPMN Analyzer 2.0: Instantaneous, Comprehensible, and Fixable Control Flow Analysis for Realistic BPMN Models

### 2024-08-01 paper.tex
Final adjustments

%% hf: enable header and footer.
\documentclass[
onecolumn, % TODO: has to be one col
% hf, % <-- Not sure if this should be enabled
]{ceurart}

@@

% This was used to mitigate some warnings
\usepackage[T1]{fontenc}
\usepackage{lmodern}
% Modern tables + extra stuff workaround
\selectcolormodel{natural}
\usepackage{ninecolors}
\selectcolormodel{rgb}
\usepackage{tabularray}

@@

% Solution
In this paper, we describe a new tool, the \textit{BPMN Analyzer 2.0}\footnote{
In the following, we will use BPMN Analyzer to refer to the BPMN Analyzer 2.0.}, for analyzing BPMN process models to detect control flow errors \textit{already} during modeling.
\autoref{fig:overview} shows an overview of the tool.
The UI is based on the popular \textit{bpmn.io} ecosystem, while the analysis is implemented in \textit{Rust} for optimal performance and memory efficiency.
We perform a breadth-first state space exploration during which we check soundness and safeness~\cite{corradiniClassificationBPMNCollaborations2018} \textit{on the fly} to uncover control flow errors.
Consequently, the tool can detect deadlocks, livelocks, starvation, dead activities, and lack of synchronization in BPMN models.
The BPMN Analyzer is open source and accessible online\footnote{\url{https://timkraeuter.com/bpmn-analyzer-js/}} alongside a video demonstration\footnote{\url{https://www.youtube.com/watch?v=Nv2W-hXNZYA}}~\cite{krauterInstantaneousComprehensibleFixable2024}.

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.5\linewidth]{images/overview}
	\caption{Overview of the BPMN Analyzer 2.0}
	\label{fig:overview}
\end{figure}

@@

\textbf{Second}, we benchmarked the tool against a synthetic data set of models that led to a state space explosion.
This represents a \textit{worst-case} scenario for formal analysis.
We generated a data set of models~\cite{krauterInstantaneousComprehensibleFixable2024} with a growing number of parallel branches with increasing length, like~\cite{corradiniFormalApproachAnalysis2021}.
\autoref{tab:parallel-branches-benchmark} shows the average runtime of our tool when analyzing these models.
The BPMN Analyzer explores the entire state space while simultaneously analyzing the control flow, i.e., verifying soundness properties.
The models' state space grows exponentially, leading to the same order of growth in runtime.
Our analysis is not instantaneous anymore when approaching 17 parallel branches of length 1 (see \autoref{tab:parallel-branches-benchmark}).
However, analysis is still instantaneous for more reasonable models with five parallel branches of length 5 or 3 branches of length 20.
Other tools report 2-3s of runtime for most soundness properties and 30s for a model with five parallel branches~\cite{corradiniFormalApproachAnalysis2021}, which took milliseconds in our tool.

\begin{table}[ht]
\scalebox{0.9}{ % Resize to 90%
    \begin{minipage}{.51\linewidth}
	   \centering
	   \caption{Benchmark results of the parallel branches models}
	   \label{tab:parallel-branches-benchmark}
	   \SetTblrInner{colsep=2pt}
	   \begin{tblr}{
	   		column{1-X} = {c},
	   		column{Y-Z} = {r},
	   		hline{1, 2, Z} = {-}{1.2pt, solid}, % Z is the last row/column
	   		hline{8, 9} = {-}{dashed},
	   		vline{2-Y} = {2-Z}{solid}, % Y is the second to last row/column
	   	}
	       \textbf{Branches} & \textbf{Branch Length} &\textbf{Runtime} & \textbf{States} \\
	       5 & 1 & 1 ms& 35 \\
	       10 & 1 & 3 ms& 1.027 \\
	       15 & 1 & 161 ms& 32.771 \\
	       16 & 1 & 360 ms& 65.539 \\
	       17 & 1 & 790 ms& 131.075 \\
	       20 & 1 & 8.803 ms& 1.048.579 \\
	       5 & 5 & 14 ms& 7.779 \\
	       3 & 20 & 11 ms& 9.264 \\
	   \end{tblr}
    \end{minipage}
    % Add extra space in the middle
    \qquad
    \qquad
    \begin{minipage}{.5\linewidth}
	   \centering
	   \caption{Benchmark results of the realistic BPMN models}
	   \label{tab:realistic-models-benchmark}
	   \SetTblrInner{colsep=2pt}
	   \begin{tblr}{
	   		column{Y-Z} = {r},
	   		hline{1, 2, Z} = {-}{1.2pt, solid}, % Z is the last row/column
	   		hline{5} = {-}{dashed},
	   		vline{2-Y} = {2-Z}{solid}, % Y is the second to last row/column
	   	}
	   	   \textbf{Model name} &\textbf{Runtime} & \textbf{States} \\
	   	   e001~\cite{houhouFirstOrderLogicVerification2022} & 1 ms & 39 \\
	   	   e002~\cite{houhouFirstOrderLogicVerification2022} & 1 ms & 39 \\
	   	   e020~\cite{houhouFirstOrderLogicVerification2022} & 10 ms & 5356 \\
	   	   credit-scoring-async\footref{footnote:camundaResearch} & 1 ms & 60 \\
	   	   credit-scoring-sync\footref{footnote:camundaResearch} & 1 ms & 140 \\
	   	   dispatch-of-goods\footref{footnote:camundaResearch} & 1 ms & 103\\
	   	   recourse\footref{footnote:camundaResearch} & 1 ms & 77 \\
	   	   self-service-restaurant\footref{footnote:camundaResearch} & 1 ms & 190 \\
	   \end{tblr}
    \end{minipage} 
}
\end{table}

@@

\section{Conclusion} \label{sec:conclusion}

@@

>>>>

%% hf: enable header and footer.
\documentclass[
onecolumn,
]{ceurart}

@@

% Modern tables + extra stuff workaround
\selectcolormodel{natural}
\usepackage{ninecolors}
\selectcolormodel{rgb}
\usepackage{tabularray}

@@

% Solution
In this paper, we describe a new tool, the \textit{BPMN Analyzer 2.0}\footnote{
In the following, we will use BPMN Analyzer to refer to the BPMN Analyzer 2.0.}, for analyzing BPMN process models to detect control flow errors \textit{already} during modeling.
\autoref{fig:overview} shows an overview of the tool.
The UI is based on the popular \textit{bpmn.io} ecosystem, while the analysis is implemented in \textit{Rust} for optimal performance and memory efficiency.
We perform a breadth-first state space exploration during which we check soundness and safeness~\cite{corradiniClassificationBPMNCollaborations2018} \textit{on the fly} to uncover control flow errors.
Consequently, the tool can detect deadlocks, livelocks, starvation, dead activities, and lack of synchronization in BPMN models.
The BPMN Analyzer is open source and accessible online alongside a video demonstration\footnote{Tool: \url{https://timkraeuter.com/bpmn-analyzer-js/}, Video: \url{https://www.youtube.com/watch?v=Nv2W-hXNZYA}}~\cite{krauterInstantaneousComprehensibleFixable2024}.

\begin{figure}[ht]
	\centering
	\includegraphics[width=0.6\linewidth]{images/overview}
	\caption{Overview of the BPMN Analyzer 2.0}
	\label{fig:overview}
\end{figure}

@@

\textbf{Second}, we benchmarked the tool against a synthetic data set of models that led to a state space explosion.
This represents a \textit{worst-case} scenario for formal analysis.
We generated a data set of models~\cite{krauterInstantaneousComprehensibleFixable2024} with a growing number of parallel branches with increasing length, like~\cite{corradiniFormalApproachAnalysis2021}.

\autoref{tab:parallel-branches-benchmark} shows the average runtime of our tool when analyzing these models.
The BPMN Analyzer explores the entire state space while simultaneously analyzing the control flow, i.e., verifying soundness properties.
The models' state space grows exponentially, leading to the same order of growth in runtime.
Our analysis is not instantaneous anymore when approaching 17 parallel branches of length 1 (see \autoref{tab:parallel-branches-benchmark}).
However, analysis is still instantaneous for more reasonable models with five parallel branches of length 5 or 3 branches of length 20.
Other tools report 2-3s of runtime for most soundness properties and 30s for a model with five parallel branches~\cite{corradiniFormalApproachAnalysis2021}, which took milliseconds in our tool.

\begin{table}[ht]
\scalebox{1}{ % Somehow needed
    \begin{minipage}{.5\linewidth}
	   \centering
	   \caption{Benchmark results of the parallel branches models}
	   \label{tab:parallel-branches-benchmark}
	   \SetTblrInner{colsep=2pt}
	   \begin{tblr}{
	   		column{1-X} = {c},
	   		column{Y-Z} = {r},
	   		hline{1, 2, Z} = {-}{1.2pt, solid}, % Z is the last row/column
	   		hline{8, 9} = {-}{dashed},
	   		vline{2-Y} = {2-Z}{solid}, % Y is the second to last row/column
	   	}
	       \textbf{Branches} & \textbf{Branch Length} &\textbf{Runtime} & \textbf{States} \\
	       5 & 1 & 1 ms& 35 \\
	       10 & 1 & 3 ms& 1.027 \\
	       15 & 1 & 161 ms& 32.771 \\
	       16 & 1 & 360 ms& 65.539 \\
	       17 & 1 & 790 ms& 131.075 \\
	       20 & 1 & 8.803 ms& 1.048.579 \\
	       5 & 5 & 14 ms& 7.779 \\
	       3 & 20 & 11 ms& 9.264 \\
	   \end{tblr}
    \end{minipage}
    % Add extra space in the middle
    \begin{minipage}{.5\linewidth}
	   \centering
	   \caption{Benchmark results of the realistic BPMN models}
	   \label{tab:realistic-models-benchmark}
	   \SetTblrInner{colsep=2pt}
	   \begin{tblr}{
	   		column{Y-Z} = {r},
	   		hline{1, 2, Z} = {-}{1.2pt, solid}, % Z is the last row/column
	   		hline{5} = {-}{dashed},
	   		vline{2-Y} = {2-Z}{solid}, % Y is the second to last row/column
	   	}
	   	   \textbf{Model name} &\textbf{Runtime} & \textbf{States} \\
	   	   e001~\cite{houhouFirstOrderLogicVerification2022} & 1 ms & 39 \\
	   	   e002~\cite{houhouFirstOrderLogicVerification2022} & 1 ms & 39 \\
	   	   e020~\cite{houhouFirstOrderLogicVerification2022} & 10 ms & 5356 \\
	   	   credit-scoring-async\footref{footnote:camundaResearch} & 1 ms & 60 \\
	   	   credit-scoring-sync\footref{footnote:camundaResearch} & 1 ms & 140 \\
	   	   dispatch-of-goods\footref{footnote:camundaResearch} & 1 ms & 103\\
	   	   recourse\footref{footnote:camundaResearch} & 1 ms & 77 \\
	   	   self-service-restaurant\footref{footnote:camundaResearch} & 1 ms & 190 \\
	   \end{tblr}
    \end{minipage} 
}
\end{table}

@@

\section{Conclusion \& Future Work} \label{sec:conclusion}

@@

In future work, we aim to improve our tool by providing more quick fixes, considering advanced BPMN elements such as different events, and ranking quick fixes based on usefulness and previous user behavior.
Finally, we aspire to test our tool in a real-world scenario to gather feedback and measure its impact on productivity.
